### 2020-12-10 submission/manuscript.Rmd
Final edits

@@

# Removal of rare sequences from 16S rRNA gene sequence surveys biases the interpretation of community structure data

@@

**word choice: sample vs. community**
**word choice: dataset vs. sample**

@@

Methods for remediating PCR and sequencing artifacts in 16S rRNA gene sequence collections are in continuous development. The choices that researchers make have significant ramifications on the inferences that can be drawn from microbial ecology studies. One common approach is to remove rare sequences from datasets. But, the threshold that is applied is generally selected without regard for the number of sequences in the sample or the variation in sequencing depth across samples within a study. I analyzed the impact of removing rare sequences on metrics of alpha and beta diversity using samples collected across `r nrow(table_1)` published datasets. Removal of rare sequences significantly decreased the number of amplicon sequence variants and operational taxonomic units as well as their diversity. Furthermore, their removal increased the variation in community structure between samples. When a known effect size was simulated, removal of rare sequences reduced the statistical power to detect the effect relative to not removing rare sequences. Although removal of rare sequences did not affect the probability of falsely detetecting differences when samples were randomized to simulate a null model, the false detection rate increased when samples were assigned to simulated treatment groups according to their sequencing depth and rare sequences were removed. The false detection rate did not vary when rare sequences were retained. This analysis demonstrates the problems inherent in removing rare sequences. Researchers are encouraged to retain these sequences, to select approaches that minimize PCR and sequencing artifacts, and to use rarefaction to control for uneven sampling effort.

@@

Removing rare sequences from 16S rRNA gene sequence collections is an approach that has grown in popularity for handling PCR and sequencing artifacts. Yet, it is unclear what impact an abundance-based filter has on downstream analyses. To investigate the effects of removing rare sequences, I analyzed the community distributions found in the samples of `r nrow(table_1)` published datasets. Analysis of these data and simulations based on them showed that removal of rare sequences distorts the representation of microbial communities. This has the effect of artificially making it more difficult to detect differences between treatment groups. Also of concern was the observation that if sequencing depth is confounded with treatment group, then the probability of falsely detecting a difference between the treatment groups increases with the removal of rare sequences. The practice of removing rare sequences should stop, lest researcher adversely affect the interpretation of their data.

@@

Numerous studies have attempted to address the problem of removing or "denoising" artifacts from data generated using Illumina's MiSeq platform. In one approach, paired sequence reads are aligned and any discrepencies between the reads are resolved based on the difference in quality score for the position in question [@Kozich2013; @Edgar2016; @Masella2012]; quality scores are also used to curate single reads [@Bokulich2013]. In addition, a polishing step is often used to identify ideal sequecnces and their related artifacts based on the frequency of the sequences and their similarity sequences [@Kozich2013; @Edgar2016; @Amir2017]. In a second approach, the quality scores and types of errors to cluster sequences directly [@Callahan2016]. Regardless of the approach, many pipelines advocate for abundance-based sceeening where rare sequences are removed from each dataset prior to outputting the sequence data as ASVs [@Edgar2016; @Amir2017; @Callahan2016]. Some algorithms recommend removing all ASVs that appear one (i.e. singletons) [@Callahan2016], eight [@Edgar2016], or ten [@Amir2017] or fewer times; these pipelines also vary in whether the minimum abundance threshold should be applied to individual samples [@Callahan2016; @Amir2017] or the pool of samples in a study [@Edgar2016]. A notable exception, the mothur-based pipeline discourages the practice of removing rare sequences [@Kozich2013]. After assigning sequences to ASVs, they are often analyzed as a taxonomic unit or clustered to generate operational taxonomic units (OTUs) or phylotypes.

@@

The minimum abundance thresholds that have been proscribed were developed and applied without regard for the number of sequences generated from each sample.  These recommendations appear to assume that the number of reads per sample is consistent within and between studies. However, it is common that the number of sequences generated from each sample may vary by two or three orders of magnitude (e.g. Figure S1). An ASV that appears once in a sample with 2,000 sequences is more trustworthy than an ASV that appears once in a sample with 100,000 sequences since it has a 50-fold higher relative abundance. But, according to the pipeline recommendations, they are treated as being equally trustworthy. Rather than removing rare ASVs, the approach taken by the mothur pipeline applies the classical ecological approach of rarefaction. Each sample is rarefied to the same sequencing depth so that the number of artifacts that appears in each sample is controlled.

Experience sequencing biological samples also demonstrates that there are bona fide ASVs that may have an abundance below the proscribed threshold. For example, the abundance of an ASV may be below the threshold in some samples or time points and above the threshold in others. However, rarity, both in terms of prevalence and incidence, is an important ecological concept. Removing rare ASVs likely hinders one's to ability to make inferences about the dynamics and nature of the populations that rare ASVs represent. Furthermore, removing ASVs whose abundances are below the proscribed threshold also potentially biases the community structure of the samples.

@@

**Datasets.** I collected `r nrow(table_1)` publicly available datasets that used the Illumina MiSeq platform to sequence the V4 region of the 16S rRNA gene from a variety of environments (Table 1). After removing poor quality and chimeric ASVs and samples that had uncharacteristically low number of sequences for the dataset, these datasets included between `r oxford_comma(n_sample_range)` samples (Figure S1). The median number of sequences for each dataset ranged between `r oxford_comma(n_med_sequence_range)` (Table 1). Strikingly, aside from the relatively small `r oxford_comma(low_fold_samples)` datasets, the difference between the sample with the fewest sequences and the sample with the most sequences for each dataset varied by between `r oxford_comma(high_fold_range, digits=1)`-fold (Table 1).

@@

**Removing treatment group effects from community data.** Because treatment effects often affect a sample's diversity and inter-sample variation, I generated null distributions for each study by randomizing, without replacement, the number of times each ASV was observed in each sample such that the total number of sequences in each sample and the total number of times each ASV was observed across all samples in the study was the same as was originally observed. This effectively made every community in a study a statistical sample of the study-wide composite community distribution. For example, after this procedure, the 490 samples from the human dataset would be expected to have the same richness and diversity of ASVs and one would not expect to find treatment-based effects between the samples. Because of the risk of bias if only one representation of the null distribution was generated, I generated 100 randomized datasets for each study. The trends between removing rare ASVs and the richness, diversity, and information loss that were identified using the observed community community distribution data were also identified with the data from the null distribution; however, the losses were larger when using the null distribution data (Figure S3). The null distribution data were used in the remainder of the study to minimize the risk of bias.

@@

**The impact of removing rare ASVs on the ability to detect statistically significant differences between treatment groups.** To test the effect of increased inter-sample variation, I randomly assigned samples to one of two treatment groups. In the first treatment group, communities were randomly sampled from the null distribution as described above. For the second treatment group, I randomly selected 10% of the ASVs in the pooled study distribution to increase their abundance by 5%. I randomly generated 100 simulated sets of treatment groups and samples. I then tested the ability to detect a difference between the two treatment groups using alpha and beta diversity metrics. The fraction of significant tests was a measurement of the statistical power to detect the difference between the treatment groups. When considering the differences in richness and diversity, the marine dataset yielded no simulated sets that were statistically significant, which was likely due to the small number of samples in the study (N=7). Among the remaining datasets, the power to detect a difference in the richness of ASVs ranged between `r oxford_comma(pull(b_base_pc_sobs,'1'), 2)` (`r oxford_comma(pull(b_base_pc_sobs,dataset))`) and between `r oxford_comma(pull(b_base_pc_shannon, '1'), 2)` (`r oxford_comma(pull(b_base_pc_shannon, dataset))`) to detect a difference in diversity when using a Wilcox test (Figure 4A). When singleton ASVs were removed, the power to detect a difference in the diversity of ASVs dropped by between `r oxford_comma(pull(b_pc_sobs, percentage_drop), 1)`% (`r oxford_comma(pull(b_pc_sobs, dataset))`) and by between `r oxford_comma(pull(b_pc_shannon, percentage_drop), 1)`% (`r oxford_comma(pull(b_pc_shannon, dataset))`; Figure 4B). The effect of removing rare ASVs on the richness of OTUs and their diversity was similar (Figure S5AB). I used the Bray-Curtis dissimilarity index to compare the simulated communities within each dataset and calculated the power to detect differences between the two simulated treatment groups using the analysis of molecular variance (Figure 4C and S5C). Without removing rare sequences, the power to detect a difference between the two simulated treatment groups varied between `r oxford_comma(pull(b_base_beta, '1') %>% range(), 2)` (`r slice(b_base_beta, 1) %>% pull(dataset)` and `r oxford_comma(slice_max(b_base_beta, '1') %>% pull(dataset))`). Aside from the `r oxford_comma(unfazed_singletons)` datasets, the power to detect differences dropped by between `r oxford_comma(pull(fazed_singletons,p_drop_singletons), 1)`% (`r oxford_comma(pull(fazed_singletons, dataset))`) when singletons were removed. However, when ASVs that occurred 10 or fewer times were removed from each sample, the power to detect differences dropped by `r oxford_comma(pull(fazed_tentons, p_drop_tens), 1)`% (`r oxford_comma(pull(fazed_tentons, dataset))`); similar results were observed when ASVs were clustered into OTUs. Removing rare ASVs reduced the ability to detect simulated treatment effects using metrics commonly used to compare microbial communities.

@@

The practice of removing rare sequences from samples seems to be a response to researchers prioritizing the number of reads and length of sequences over their quality [@Bokulich2013]. Previous work has shown that assembly of fully overlapping sequence reads results in the lowest sequencing error rates [@Kozich2013]. The studies highlighted in this analysis sequenced the V4 region of the 16S rRNA gene using the Illumina MiSeq sequencing platform with their v2 chemistry. The resulting data consists of two 250 nt reads that span a region that is about 250 nt long. In contrast, there has been a movement to sequence longer regions with similar chemistry resulting in less overlap between the sequencing reads [@Fadrosh2014; @Holm2019]. Alternatively, others have prioritized increasing the number of sequences per sample by sequencing the V4 region, but with paired or single reads that are 150 nt long [@Thompson2017; @McDonald2018]. Both practices result in a significantly higher error rate for the resulting assmebly. Instead, researchers should prioritize the quality over the quantity and length of their data. For these reasons, this analysis did not use lower quality data generated by alternative methods

The impacts of removing rare sequences on the representation of communities are caused by the uneven impact of applying a single abundance threshold across all samples. The number of sequences per sample in the datasets highlighted in this analysis varied by 100-fold (Figure S1); I suspect the range is actually larger since researchers may have opted against depositing samples with fewer reads into the SRA. If the range in read coverage for a study is 100-fold, a singleton in a small sample would have had a comparable relative abundance as a sequence that appeared 100 times in a more densely sequenced sample. However, it would have been removed from the smaller sample and not the larger sample. Thus, applying a single abundance threshold disproportionately impacted the samples with fewer sequences. This seems to contradict the purpose of removing rare sequences since a singleton in the smaller sample is more reliable than the singleton in the larger sample. A superior approach to removing rare sequences is to use rarefaction to conrol for uneven sampling.

@@

**Data curation and analysis.** To insure the highest possible data quality, datasets were limited to those where the 500 cycle v2 MiSeq chemistry was used to sequence the amplicons. The paired 250 nt reads resulted in near complete 2-fold sequencing coverage of every nucleotide in the ca. 250 nt-long region. This region and sequencing platform were selected because previous work has shown that a standard data analysis pipeline in mothur results in a sequencing error rate below 0.02% [@Kozich2013]. All sequence data were obtained from the Sequence Read Archive (SRA) and processed using a standard mothur-based sequencing pipeline that resulted in ASVs as generated by the pre.cluster algorithm using a threshold of 2 nt [@Kozich2013; @Schloss2009]. ASVs were assigned to OTUs using a 3% distance threshold using mothur's cluster function with the OptiClust algorithm [@Westcott2017]. To minimize the effects of uneven sampling effort, samples were rarefied to the number of sequences in the smallest sample for each dataset. Because metrics of alpha diversity did not consistently follow a normal distribution, I used the non-parametric Wilcoxon rank test as implemented in R. For comparisons of Bray-Curtis distances, the amova function within mothur was used, which implements the analysis of molecular variance algorithm [@Anderson2001].

**Reproducibility.** All analyses were performed using mothur (version 1.44.1) and R (version `r paste(R.version$major, R.version$minor, sep='.')`) with the tidyverse (v.`r packageVersion("tidyverse")`), broom (v.`r packageVersion("broom")`), data.table (v.`r packageVersion("data.table")`), and cowplot (v.`r packageVersion("cowplot")`) packages. All of the code that was used in this analysis as well as the Makefile for running the analysis are available at https://github.com/SchlossLab/Schloss_Singletons_XXXXX_2019 along with the complete version history of the project.

>>>>

package_version <- function(package){

  paste(unlist(packageVersion(package)), collapse='.')

}

@@

# Removal of rare amplicon sequence variants from 16S rRNA gene sequence surveys biases the interpretation of community structure data

@@

@@

Methods for remediating PCR and sequencing artifacts in 16S rRNA gene sequence collections are in continuous development and have significant ramifications on the inferences that can be drawn. A common approach is to remove rare amplcon sequence variants (ASVs) from datasets. But, the definition of rarity is generally selected without regard for the number of sequences in the samples or the variation in sequencing depth across samples within a study. I analyzed the impact of removing rare ASVs on metrics of alpha and beta diversity using samples collected across `r nrow(table_1)` published datasets. Removal of rare ASVs significantly decreased the number of ASVs and operational taxonomic units as well as their diversity. Furthermore, their removal increased the variation in community structure between samples. When simulating a known effect size, removal of rare ASVs reduced the power to detect the effect relative to not removing rare ASVs. Removal of rare ASVs did not affect the false detection rate when samples were randomized to simulate a null model. However, the false detection rate increased when rare ASVs were removed using a null distribution and assignment of samples to simulated treatment groups according to their sequencing depth. The false detection rate did not vary when rare ASVs were retained. This analysis demonstrates the problems inherent in removing rare ASVs. Researchers are encouraged to retain rare ASVs, to select approaches that minimize PCR and sequencing artifacts, and to use rarefaction to control for uneven sequencing effort.

@@

Removing rare amplicon sequence variants (ASVs) from 16S rRNA gene sequence collections is an approach that has grown in popularity for limiting PCR and sequencing artifacts. Yet, it is unclear what impact an abundance-based filter has on downstream analyses. To investigate the effects of removing rare ASVs, I analyzed the community distributions found in the samples of `r nrow(table_1)` published datasets. Analysis of these data and simulations based on them showed that removal of rare ASVs distorts the representation of microbial communities. This has the effect of artificially making it more difficult to detect differences between treatment groups. Also of concern was the observation that if sequencing depth is confounded with the treatment, then the probability of falsely detecting a difference between the treatment groups increased with the removal of rare ASVs. The practice of removing rare ASVs should stop, lest researcher adversely affect the interpretation of their data.

@@

Numerous studies have attempted to address the problem of removing or "denoising" artifacts from data generated using Illumina's MiSeq platform. In one approach, paired sequence reads are aligned and any discrepencies between the reads are resolved based on the difference in quality score for the position in question [@Kozich2013; @Edgar2016; @Masella2012]; quality scores are also used to curate single reads [@Bokulich2013]. In addition, a polishing step is often used to identify ASVs based on the frequency and similarity of sequences [@Kozich2013; @Edgar2016; @Amir2017]. In a second approach, the quality scores and types of errors are modelled to cluster sequence reads directly into ASVs [@Callahan2016]. Regardless of the approach, many pipelines advocate for abundance-based screening where rare sequences are removed from each dataset prior to outputting the sequence data as ASVs [@Edgar2016; @Amir2017; @Callahan2016]. Some algorithms recommend removing all ASVs that appear one (i.e. singletons) [@Callahan2016], eight [@Edgar2016], or ten [@Amir2017] or fewer times; these pipelines also vary in whether the minimum abundance threshold should be applied to individual samples [@Callahan2016; @Amir2017] or the pool of samples in a study [@Edgar2016]. A notable exception, the mothur-based pipeline discourages the practice of removing rare sequences [@Kozich2013]. After assigning sequences to ASVs, ASVs are often analyzed as a taxonomic unit or clustered to generate operational taxonomic units (OTUs) or phylotypes.

@@

The minimum abundance thresholds that have been proscribed were developed and applied without regard for the number of sequences generated from each sample.  These recommendations appear to assume that the number of reads per sample is consistent within and between studies. However, it is common that the number of sequences generated from each sample may vary by two or three orders of magnitude (e.g. Table 1 and Figure S1). An ASV that appears once in a sample with 2,000 sequences is more trustworthy than an ASV that appears once in a sample with 100,000 sequences since it has a 50-fold higher relative abundance. But, according to the pipeline recommendations, they are treated as being equally trustworthy. Rather than removing rare ASVs, the approach taken by the mothur pipeline applies the classical ecological approach of rarefaction. Each sample is rarefied to the same sequencing depth so that the number of artifacts that appears in each sample is controlled.

Experience sequencing biological samples also demonstrates that there are bona fide ASVs that may have an abundance below the proscribed threshold. For example, the abundance of an ASV may be below the threshold in some samples or time points and above the threshold in others. However, rarity, both in terms of prevalence and incidence, is an important ecological concept [@Reid2011]. Removing rare ASVs likely hinders one's to ability to make inferences about the dynamics and nature of the populations that rare ASVs represent. Furthermore, removing ASVs whose abundances are below the proscribed threshold also potentially biases the community structure of the samples.

@@

**Datasets.** I collected `r nrow(table_1)` publicly available datasets that used the Illumina MiSeq platform to sequence the V4 region of the 16S rRNA gene from a variety of environments (Table 1; Figure S1). After removing poor quality and chimeric ASVs and samples that had uncharacteristically low number of sequences for the dataset, these datasets included between `r oxford_comma(n_sample_range)` samples. The median number of sequences for each dataset ranged between `r oxford_comma(n_med_sequence_range)`. Strikingly, aside from the relatively small `r oxford_comma(low_fold_samples)` datasets, the difference between the sample with the fewest sequences and the sample with the most sequences for each dataset varied by between `r oxford_comma(high_fold_range, digits=1)`-fold.

@@

**Removing treatment group effects from community data.** Because treatment effects often affect a sample's diversity and inter-sample variation, I generated null distributions for each study by randomizing, without replacement, the number of times each ASV was observed in each sample such that the total number of sequences in each sample and the total number of times each ASV was observed across all samples in the study was the same as was originally observed. This effectively made every community in a study a statistical sample of the study-wide composite community distribution. For example, after this procedure, each of the 490 samples from the human dataset would be expected to have the same richness and diversity of ASVs and one would not expect to find treatment-based effects between the samples. Because of the risk of bias if only one representation of the null distribution was generated, I generated 100 randomized datasets for each study. The trends between removing rare ASVs and the richness, diversity, and information loss that were identified using the observed community distribution data were also identified with the data from the null distribution; however, the losses were larger when using the null distribution data (Figure S3). The null distribution data were used in the remainder of the study to minimize the risk of bias.

@@

**The impact of removing rare ASVs on the ability to detect statistically significant differences between treatment groups.** To test the effect of increased inter-sample variation, I randomly assigned samples to one of two treatment groups. In the first treatment group, communities were randomly sampled from the null distribution as described above. For the second treatment group, I randomly selected 10% of the ASVs in the pooled study distribution to increase their abundance by 5%. I randomly generated 100 simulated sets of treatment groups and samples. I then tested the ability to detect a difference between the two treatment groups using alpha and beta diversity metrics. The fraction of significant tests was a measurement of the statistical power to detect the difference between the treatment groups. When considering the differences in richness and diversity, the marine dataset yielded no simulated sets that were statistically significant, which was likely due to the small number of samples in the study (N=7). Among the remaining datasets, the power to detect a difference in the richness of ASVs ranged between `r oxford_comma(pull(b_base_pc_sobs,'1'), 2)` (`r oxford_comma(pull(b_base_pc_sobs,dataset))`) and between `r oxford_comma(pull(b_base_pc_shannon, '1'), 2)` (`r oxford_comma(pull(b_base_pc_shannon, dataset))`) to detect a difference in diversity when using a Wilcox test (Figure 4A). When singleton ASVs were removed, the power to detect a difference in the richness of ASVs dropped by between `r oxford_comma(pull(b_pc_sobs, percentage_drop), 1)`% (`r oxford_comma(pull(b_pc_sobs, dataset))`) and by between `r oxford_comma(pull(b_pc_shannon, percentage_drop), 1)`% for their diversity(`r oxford_comma(pull(b_pc_shannon, dataset))`; Figure 4B). The effect of removing rare ASVs on the richness of OTUs and their diversity was similar (Figure S5AB). I used the Bray-Curtis dissimilarity index to compare the simulated communities within each dataset and calculated the power to detect differences between the two simulated treatment groups using the analysis of molecular variance (Figure 4C and S5C). Without removing rare sequences, the power to detect a difference between the two simulated treatment groups varied between `r oxford_comma(pull(b_base_beta, '1') %>% range(), 2)` (`r slice(b_base_beta, 1) %>% pull(dataset)` and `r oxford_comma(slice_max(b_base_beta, '1') %>% pull(dataset))`). Aside from the `r oxford_comma(unfazed_singletons)` datasets, the power to detect differences dropped by between `r oxford_comma(pull(fazed_singletons,p_drop_singletons), 1)`% (`r oxford_comma(pull(fazed_singletons, dataset))`) when singletons were removed. However, when ASVs that occurred 10 or fewer times were removed from each sample, the power to detect differences dropped by `r oxford_comma(pull(fazed_tentons, p_drop_tens), 1)`% (`r oxford_comma(pull(fazed_tentons, dataset))`); similar results were observed when ASVs were clustered into OTUs. Removing rare ASVs reduced the ability to detect simulated treatment effects using metrics commonly used to compare microbial communities.

@@

The practice of removing rare sequences from samples seems to be a response to researchers prioritizing the number of reads and length of sequences over their quality [@Bokulich2013]. Previous work has shown that assembly of fully overlapping sequence reads results in the lowest sequencing error rates [@Kozich2013]. The studies highlighted in this analysis sequenced the V4 region of the 16S rRNA gene using the Illumina MiSeq sequencing platform with their version 2 chemistry. The resulting data consists of two 250 nt reads that span a region that is about 253 nt long. In contrast, there has been a movement to sequence longer regions with similar chemistry resulting in less overlap between the sequencing reads [@Fadrosh2014; @Holm2019]. Alternatively, others have prioritized increasing the number of sequences per sample by sequencing the V4 region, but with paired or single reads that are 150 nt long [@Thompson2017; @McDonald2018]. Both practices result in a significantly higher error rate for the resulting assmebly. Instead, researchers should prioritize the quality over the quantity and length of their data. For these reasons, this analysis did not use lower quality data generated by alternative methods.

The impacts of removing rare sequences on the representation of communities are caused by the uneven impact of applying a single abundance threshold across all samples. The number of sequences per sample in the datasets highlighted in this analysis varied by as much as 100-fold (Table 1; Figure S1). In practice, I suspect this range is actually larger since researchers may have opted against depositing samples with fewer reads into the SRA. If the range in read coverage for a study is 100-fold, a singleton in a small sample would have had a comparable relative abundance as a sequence that appeared 100 times in a more densely sequenced sample. However, it would have been removed from the smaller sample and not the larger sample. Thus, applying a single abundance threshold disproportionately impacted the samples with fewer sequences. This seems to contradict the purpose of removing rare sequences since a singleton in the smaller sample is more reliable than the singleton in the larger sample. A superior approach to removing rare sequences is to use rarefaction to conrol for uneven sampling.

@@

**Data curation and analysis.** To insure the highest possible data quality, datasets were limited to those where the 500 cycle version 2 MiSeq chemistry was used to sequence the amplicons. The paired 250 nt reads resulted in near complete 2-fold sequencing coverage of every nucleotide in the ca. 253 nt-long region. This region and sequencing platform were selected because previous work has shown that a standard data analysis pipeline in mothur results in a sequencing error rate below 0.02% [@Kozich2013]. All sequence data were obtained from the Sequence Read Archive (SRA) and processed using a standard mothur-based sequencing pipeline that resulted in ASVs as generated by the pre.cluster algorithm using a threshold of 2 nt [@Kozich2013; @Schloss2009]. ASVs were assigned to OTUs using a 3% distance threshold using mothur's cluster function with the OptiClust algorithm [@Westcott2017]. To minimize the effects of uneven sampling effort, samples were rarefied to the number of sequences in the smallest sample for each dataset. Because metrics of alpha diversity did not consistently follow a normal distribution, I used the non-parametric Wilcoxon rank test as implemented in R. For comparisons of Bray-Curtis distances, the amova function within mothur was used, which implements the analysis of molecular variance algorithm [@Anderson2001].

**Reproducibility.** All analyses were performed using mothur (version 1.44.1) and R (version `r paste(R.version$major, R.version$minor, sep='.')`) with the tidyverse (version `r package_version("tidyverse")`), broom (version `r package_version("broom")`), data.table (version `r package_version("data.table")`), and cowplot (version `r package_version("cowplot")`) packages. All of the code that was used in this analysis as well as the Makefile for running the analysis are available at https://github.com/SchlossLab/Schloss_Singletons_XXXXX_2019 along with the complete version history of the project.
